### 2013-08-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
moving articial absorption into an appendix and the reduced domain approximation into the spn chapter

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a brief variance
analysis of several Monte Carlo estimators. The Sequential Monte Carlo
method of Halton and Monte Carlo Synthetic Acceleration methods will
then be presented as a means of leveraging the basic Monte Carlo
methods devised by Neumann and Ulam in an iterative refinement scheme
that accelerates their convergence. Using these iterative schemes, a
set of modest variance reduction techniques will be introduced and
their effects along with other parameters of the algorithms on the
Monte Carlo solutions explored using a simple model transport problem.

@@

%%---------------------------------------------------------------------------%%
\section{Variance Reduction Techniques}
\label{sec:variance_reduction}
Variance reduction is a mechanism by which the probabilities and
weights of the Monte Carlo problem are modified to improve the
performance of a given estimator. In the context of improving MCSA
results, such modifications may potentially introduce bias into the
system. However, using Monte Carlo within an MCSA iterative scheme
provides a potential buffer for the solution from this bias as the
correction generated by the Monte Carlo solve will contain large
statistical error even without variance reduction. For this work, any
method that alters the weights and probabilities of the Monte Carlo
method with the aim of improving either iterative performance or time
to convergence will be defined as a variance reduction scheme.

\subsection{Artificial Absorption}
\label{subsec:artificial_absorption}
The random walk sequences generated by the adjoint Neumann Ulam
decomposition in Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_weight}) will continue on indefinitely without the
implementation of a weight cutoff procedure (which in itself is
effectively a form of biased variance reduction). This is due to the
fact that all states to which a stochastic history may move in a given
transition exist within the system and have a non-zero weight. A
preliminary variance reduction scheme is to introduce
\textit{artificial absorption} into the system such that at each
transition event, a stochastic history will now have some finite
probability of being terminated through absorption as well as
transitioning to other states in the system. This probability,
$p_{abs}$, modifies the probabilities and weights in the adjoint
random walk sequence as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j|h_{ji}|}(1-p_{abs})\:,
  \label{eq:absorption_probabilities}
\end{equation}
and
\begin{equation}
  w_{ij} = \frac{sign(h_{ji})}{(1-p_{abs})}\sum_j|h_{ji}|\:.
  \label{eq:absorption_weights}
\end{equation}
At each transition step, a history at state $i$ will then sample the
probability distribution function generated by
Eq~(\ref{eq:absorption_probabilities}) with an additional absorption
state added to the distribution. If the sampling procedure results in
a transition to the absorption state, the history is immediately
terminated. The advantages of this for residual Monte Carlo are that
increasing the absorption probability decreases the length of each
random walk (resulting in speedup), while the random walk has a higher
weight contribution at every step. This higher weight means that each
history is depositing more information near its birth site which will
on average be located where the system residual is the largest. An
additional advantage of this formulation is that the full Neumann-Ulam
decomposition is maintained.

To observe the effects of this variance reduction on MCSA solutions,
the transient Poisson problem was again solved with the adjoint
method, this time with varying values of artificial absorption. A $200
\times 200$ grid was used with 40,000 histories at every iteration
with the expected value estimator converged to a tolerance of
\sn{1}{-8}. To reduce the effect of the weight cutoff on this study,
the weight cutoff was set to \sn{1}{-12} such that it is significantly
more likely that a history will be terminated by absorption rather
than weight cutoff. Figure~\ref{fig:absorption_iters} gives the number
of iterations to converge as a function of the artificial absorption
probability. Up to a probability of roughly 0.5, the iterative
performance is not significantly affected with the information lost
due to shortened random walks causing a small increase in the number
of iterations. At a probability of 0.6, the number of iterations
required grows rapidly as the shortened randow walks create an MCSA
correction that has a larger statistical error. Convergence was
observed to be lost at probabilities of 0.7 and higher.

The random walk length has a strong effect on CPU time as
well. Figure~\ref{fig:absorption_time} gives the CPU time in seconds
required to converge as a function of artificial absorption
probability. Initially, small absorption probabilities not only
maintain iterative performance, but also drastically reduced the time
required to converge as absorption was more frequent than even a large
weight cutoff but the weights themselves were not significantly
modified. As the number of iterations grow rapidly, so does the time
required to converge to a solution. Based on these results, using a
small amount of artificial absorption should provide some CPU speedup
while maintaining iterative performance. For the problem presented
here, an absorption probability of around 0.25 may be the best choice
as it minimizes CPU time.

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. artificial
      absorption probability for a $200 \times 200$ square mesh and
      40,000 histories.}  \textit{The addition of absorption is
      detrimental to iterative performance due to both the shortening
      of the random walks as well as the modification of the
      transition weight.}}
  \label{fig:absorption_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
      probability for a $200 \times 200$ square mesh and 40,000
      histories.}  \textit{The reduced random walk length speeds up
      the calculation with a fixed number of histories. Too much
      artificial absorption increases iterations rapidly giving an
      increasing CPU time.}}
  \label{fig:absorption_time}
\end{figure}
\clearpage

\subsection{Reduced Domain Approximation}
\label{subsec:reduced_domain_approximation}
For scalable parallel implementations and efficient Monte Carlo
transport, the domain is required to be sparse. We may find through
general preconditioning operations as outlined by the Monte Carlo
problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
lost. To alleviate this, we propose modifying the domain sampled by
the Monte Carlo method to generate the correction within MCSA by
removing terms in the iteration matrix through some predetermined
criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
stochastic realization of multiple matrix-vector multiplies, the
criteria we choose should be based on maintaining those elements that
will make the largest contributions to the multiplication and
therefore recover as much of the original product as possible.

For each row in the preconditioned system, we can then apply the
reduced domain approximation to eliminate those elements that either
fall below a specified tolerance level, keep the $N$ largest elements
in each row, or both giving:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
    \label{eq:rda_correction}\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:rda_mcsa}
\end{subequations}
where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
reduced domain approximation has been applied to the iteration matrix
in the Monte Carlo method. If enough of the terms in the original
iteration matrix have been maintained, the character of the original
iteration matrix should be retained and continue to accelerate the
iterative scheme. The effects of the reduced domain approximation on
neutron transport systems where it is required due to the explicit
preconditioning strategy will be presented in
Chapter~\ref{ch:spn_equations}.

@@

\begin{itemize}
\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
  for all discrete linear transport problems
\item MCSA can be preconditioned using a general left/right scheme but
  requires the explicit inversion of the preconditioners and
  construction of the composite linear operator
\item The adjoint Neumann-Ulam method performs better when used with
  MCSA compared to the forward method
\item MCSA has superior performance when compared to Halton's residual
  Monte Carlo method
\item The Richardson iteration in MCSA behaves as a smoother for the
  stochastic Monte Carlo error
\item MCSA iterative performance when used with the expected value
  estimator is insensitive to histories per iteration
\item MCSA iterative performance is insensitive to the stochastic
  history weight cutoff
\item Variance reduction with artificial absorption can enhance time
  to solution
\item The reduced domain approximation was introduced to mitigate the
  consequences of composite linear operator generation
\end{itemize}

>>>>

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a brief variance
analysis of several Monte Carlo estimators. The Sequential Monte Carlo
method of Halton and Monte Carlo Synthetic Acceleration methods will
then be presented as a means of leveraging the basic Monte Carlo
methods devised by Neumann and Ulam in an iterative refinement scheme
that accelerates their convergence. Using these iterative schemes,
the effects of the algorithm parameters on solutions to a simple model
transport problem are explored.

@@

@@

\begin{itemize}
\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
  for all discrete linear transport problems
\item MCSA can be preconditioned using a general left/right scheme but
  requires the explicit inversion of the preconditioners and
  construction of the composite linear operator
\item The adjoint Neumann-Ulam method performs better when used with
  MCSA compared to the forward method
\item MCSA has superior performance when compared to Halton's residual
  Monte Carlo method
\item The Richardson iteration in MCSA behaves as a smoother for the
  stochastic Monte Carlo error
\item MCSA iterative performance when used with the expected value
  estimator is insensitive to histories per iteration
\item MCSA iterative performance is insensitive to the stochastic
  history weight cutoff
\end{itemize}
